---
title: "Mixed-effects models with R"
subtitle: "Part 4: Fitting Complex Models to Large Datasets"
author: "Douglas Bates"
date: "2018/09/20 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: left, top

```{r setup,include=FALSE}
options(width=85,show.signif.stars=FALSE,str=strOptions(strict.width="cut"))
library(lattice)
library(RePsychLing)
library(lme4)
library(JuliaCall)
lattice.options(default.theme = function() standard.theme())
library(knitr)
opts_chunk$set(prompt=TRUE,comment=NA)
``` 

# Large experimental or observational studies

- Some experiments or studies in Psychology involve collecting data on many subjects under a variety of experimental conditions, resulting in large data sets.

- The experimental design in such studies incorporates *experimental factors* (e.g. priming, location, word-length)
and *blocking factors* (e.g. subject, item)

- In modeling the response the experimental factors are incorporated as fixed effects and the blocking factors as random effects, leading to the use of mixed-effects models.

- Recent papers have advocated fitting "maximal models" in the random-effects terms, leading to difficult optimization problems

---
class: left, top

# Data from Kliegl et al.

```{r KKLdata}
str(KKL)
```

---
class: left, top
```{r m0fit}
form <- lrt ~ sze * (spt + obj + grv) * orn + (spt + obj + grv + orn + spt_orn + obj_orn + grv_orn | subj)
proc.time(m0 <- lmer(form, KKL, REML=FALSE))
```

